---
title: "Final Project"
subtitle: "Data Science for Biologists, Fall 2021"
author: "Courtney Charlesworth"
output: 
  html_document:
    theme: readable
    highlight: pygments
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# DON'T CHANGE THE NEXT LINE, except to add NEW options if you have any, like default figure sizes
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo = TRUE, include = TRUE)

# Path to figure export directory 
path_to_figures <- file.path(here::here(), "figures")

# Load all libraries you use HERE. If you have to install something, please make sure to discuss with Spielman.
library(tidyverse)
library(modelr)

#plot theme
theme_set(theme_classic())
```

<br><br>

## Introduction

This [dataset](https://archive.ics.uci.edu/ml/datasets/Forest+Fires) contains information recorded from forest fires in Portugal, such as temperature (`temp`) and relative humidity (`RH`). The goal of the researchers who put this together was to make a model that can predict how large the burned `area` of the forest will be. I chose this dataset because fire is cool... and because it would be interesting to also make a model predicting burn `area`.

The variables in the dataset are as follows (*copy and pasted from website*):

+ `X` - x-axis spatial coordinate within the Montesinho park map: 1 to 9
+ `Y` - y-axis spatial coordinate within the Montesinho park map: 2 to 9
+ `month` - month of the year: 'jan' to 'dec'
+ `day` - day of the week: 'mon' to 'sun'
+ `FFMC` - FFMC index from the FWI system: 18.7 to 96.20
+ `DMC` - DMC index from the FWI system: 1.1 to 291.3
+ `DC` - DC index from the FWI system: 7.9 to 860.6
+ `ISI` - ISI index from the FWI system: 0.0 to 56.10
+ `temp` - temperature in Celsius degrees: 2.2 to 33.30
+ `RH` - relative humidity in %: 15.0 to 100
+ `wind` - wind speed in km/h: 0.40 to 9.40
+ `rain` - outside rain in mm/m2 : 0.0 to 6.4
+ `area` - the burned area of the forest (in ha): 0.00 to 1090.84 (this output variable is very skewed towards 0.0, thus it may make sense to model with the logarithm transform).


```{r read-examine-data}
# Code to read in the data:
forest_fires <- read_csv("charlesworth_forestfires.csv")

# Code to dplyr::glimpse() the data:
dplyr::glimpse(forest_fires)

```

<br><br>

## Question 1: Modeling

#### Does `month` predict burn `area`?

### Methods

```{r question1_modelsetup}
#model selection
#linear model, hypothesis-based analysis

#lm(response ~ predictors, data = nameofthetibble)
model_area <- lm(area ~ month, data = forest_fires) 

#show model
model_formula <- as.formula(model_area)
model_formula # look at formula/the predictors

```


```{r question1_testmodel}
#R^2 and RMSE to interpret how good the model is
#model was very weak even if 0s from area were excluded (did not make a huge difference at all)
rsquare(model_area, forest_fires) #0.005477511
rmse(model_area, forest_fires) #63.41982

#helps with interpreting RMSE
summary(forest_fires$area)

```


```{r question1_tibble}
#tibble
forest_fires %>%
  filter(area < 200) %>% #remove really extreme outliers so plot isnt compressed
  select(month, area) %>% #tibble output looks nicer
  group_by(month) %>% 
  mutate(row = 1:n()) %>% #assigned so can pivot
  #pivot to set up rename()
  pivot_wider(names_from = month,
              values_from = area) %>%
  #rename months
  rename("January" = jan,
         "Febraury" = feb,
         "March" = mar,
         "April" = apr,
         "May" = may,
         "June" = jun,
         "July" = jul,
         "August" = aug,
         "September" = sep,
         "October" = oct,
         "November" = nov,
         "December" = dec) %>% 
  pivot_longer(March:last_col(), #March (first datapoint) through the last column
               names_to = "month",
               values_to = "area") -> tibble_q1

tibble_q1

```


```{r question1_plot, fig.height=6, fig.width=8}
#plotting
tibble_q1 %>%
  ggplot() +
  aes(x = fct_relevel(month, #reorder graph
                      "January",
                      "Febraury",
                      "March",
                      "April",
                      "May",
                      "June",
                      "July",
                      "August",
                      "September",
                      "October",
                      "November",
                      "December"),
      y = area) +
  geom_boxplot() +
  labs(x = "Month",
       y = "Area (ha)",
       title = "Question 1") -> plot_q1

plot_q1
```


```{r question1_savefigure, include=FALSE}
#save plot to file
plot_q1_file <- file.path(path_to_figures, #path to figures
                          "question1.png") #file name with file type (.png)

ggsave(plot_q1_file, #path to file where figure gets saved
       plot_q1, #variable with the plot/image
       width = 8, height = 6) #dimensions of the figure (inches)

```


### Answer

$R^2 = 0.0054$, which means that 99.5% of variation is unexplained, indicating that this is a weak model. The RMSE is 63.42 and the values range from 0.00 to 1090.84. This is further evidence that the model is weak as the data also skews toward 0 and there are a lot of outliers.


<br><br>

## Question 2

#### What `day` of the week has the largest forest fires?

### Methods

```{r question2_tibble}
#data wrangling
forest_fires %>%
  filter(area != 0) %>% #exclude areas that equal 0, wanted fires w/ actual area value to measure/summarize
  group_by(day) %>% 
  summarize(mean_area = mean(area)) %>%
  #pivot to set up rename()
  pivot_wider(names_from = day,
              values_from = mean_area) %>%
  #rename days of the week
  rename("Friday" = fri,
         "Monday" = mon,
         "Saturday" = sat,
         "Sunday" = sun,
         "Thursday" = thu,
         "Tuesday" = tue,
         "Wednesday" = wed) %>%
  #pivot back to set up for plotting
  pivot_longer(Friday:Wednesday,
               names_to = "day",
               values_to = "mean_area") -> q2_tibble
q2_tibble

```


```{r question2_plot, fig.height=4, fig.width=7}
#plotting
q2_tibble %>%
  ggplot() +
  aes(x = fct_relevel(day, #ordered by sequential order
                           "Monday", "Tuesday", "Wednesday", 
                           "Thursday", "Friday", "Saturday", "Sunday"),
      y = mean_area,
      fill = day) +
  geom_col() +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = "Day of the week",
       y = "Mean burn area (ha)",
       title = "Question 2") +
  #removes legend
  theme(legend.position = "none") -> plot_q2

plot_q2

```


```{r question2_savefigure, include=FALSE}
#save plot to file
plot_q2_file <- file.path(path_to_figures, #path to figures
                          "question2.png") #file name with file type (.png)

ggsave(plot_q2_file, #path to file where figure gets saved
       plot_q2, #variable with the plot/image
       width = 7, height = 4) #dimensions of the figure (inches)

```


### Answer

Saturday tends to have the largest forest fires.

<br><br>

## Question 3

#### What `season` has the most forest fires?

### Methods

```{r question3_tibble}
#data wrangling
forest_fires %>%
  #categorize months by season, new variable/column
  mutate(season = case_when(month %in% c("mar", "apr", "may") ~ "Spring",
                            month %in% c("jun", "jul", "aug") ~ "Summer",
                            month %in% c("sep", "oct", "nov")  ~ "Fall",
                            month %in% c ("dec", "jan", "feb") ~ "Winter")) %>%
  group_by(season) %>%
  count() -> tibble_q3 #count how many datapoints there are for each season

tibble_q3

```
 

```{r question3_plot}
#plotting
tibble_q3 %>%
  ggplot() +
  aes(x = fct_relevel(season,
                      "Spring", "Summer", "Fall", "Winter"),
      y = n,
      fill = season) +
  geom_col(color = "black") +
  scale_fill_manual(values = c("darkorange", #seasonal colors
                               "mediumorchid1",
                               "yellow",
                               "lightblue")) +
  labs(x = "Season",
       y = "Number of forest fires",
       title = "Question 3") +
  #removes legend
  theme(legend.position = "none") -> plot_q3

plot_q3

```


```{r question_savefigure, include=FALSE}
#save plot to file
plot_q3_file <- file.path(path_to_figures, #path to figures
                          "question3.png") #file name with file type (.png)

ggsave(plot_q3_file, #path to file where figure gets saved
       plot_q3, #variable with the plot/image
       width = 6, height = 4) #dimensions of the figure (inches)

```


### Answer

Summer tends to have the most forest fires.


<br><br>

## Question 4

#### Do hotter days tend to be more humid?

### Methods

```{r question4_tibble}
#data wrangling
forest_fires %>%
  #categorize temp as High or Low depending on if the value is above or below the mean
  mutate(temp_threshold = ifelse(temp >= mean(temp), 
                                 "High", 
                                 "Low")) %>%
  select(RH, temp_threshold) -> tibble_q4 #nicer tibble output

tibble_q4

```


```{r question4_plot_mean_or_median, include=FALSE}
#summary statistics
summary(forest_fires$temp) #mean is not resistant to outliers, while median is

#we do not need to use the median for plotting
#we can use the mean because the data is approximately normally distributed, with only a small amount of skew 
forest_fires %>%
  ggplot() +
  aes(x = temp) +
  geom_density() -> plot_q4_mean_or_median

plot_q4_mean_or_median

```


```{r question4_plot}
#plotting
tibble_q4 %>%
  ggplot() +
  aes(x = RH,
      fill = temp_threshold) +
  geom_density(alpha = 0.4) +
  scale_fill_viridis_d(option = "viridis") +
  labs(x = "Relative humidity",
       y = "Density",
       title = "Question 4",
       fill = "Temperature",
       caption = "High = above 18.89 Celsius, Low = below 18.89 Celsius") -> plot_q4

plot_q4

```


```{r question4_savefigure, include=FALSE}
#save plot to file
plot_q4_file <- file.path(path_to_figures, #path to figures
                          "question4.png") #file name with file type (.png)

ggsave(plot_q4_file, #path to file where figure gets saved
       plot_q4, #variable with the plot/image
       width = 6, height = 4) #dimensions of the figure (inches)

```


### Answer

Days with a high temperature tend to have a lower humidity than days with a low temperature. Although it should be noted that the larger spread for low temperature days may be causing the mean to shift to the right. 
